Parsing Postfix log files

### 2007-03-21 doc/logparser.tex
Most of the structural changes come from today's meeting with Carl. Added an abstract and table of contents; previous abstract became introduction. Started writing a parser background section, suggested by Carl. Added limitations and possible improvements section (partially written), and empty conclusion section. Lots of tweaks and minor improvements, re-wording, etc.

\title{Parsing Postfix log files}
\author{John Tobin \\ Computer Science Department \\ Trinity College \\
Dublin 2 \\ Ireland \\ tobinjt@cs.tcd.ie}
\date{}
\maketitle

\section{Abstract}

Most mail server administrators will have performed some basic processing
of Postfix logs at one time or another, whether it was to debug out a
problem, explain why mail is being rejected, or to check whether their new
anti-spam measures are working.  The more adventurous will have generated
some statistics to show how many hits each of their anti-spam measures have
gotten in the last week, and possibly even generated some graphs to clearly
illustrate the point to management or complaining users\footnote{This was
the author's first real foray into processing Postfix logs}.  Very few will
have performed in-depth parsing and analysis of their logs, where the
parsing must correlate the log lines per-connection or per-queueid, rather
than processing lines independently.  One of the barriers to this type of
processing is the unstructured nature of Postfix's logs, where each logging
line was added on an ad hoc basis as Wietse found a requirement or added
new functionality, and further complication comes from user defined log
messages.  This paper attempts to document the difficult process of parsing
Postfix logs, and presents a program which parses logs and places the
resulting data into a database for further processing.

@@

Postfix is a highly configurable, high performance, secure and scalable
Mail Transport Agent.  It features extensive anti-spam restrictions,
allowing a mail administrator to deploy those restrictions which they judge
suitable for their needs, rather than a fixed set chosen by Postfix's
author.  These restrictions can be applied and combined on a per-host,
per-recipient or per-sender basis, allowing clients/recipients to be
whitelisted with respect to specific restrictions.  Postfix leverages
simple lookup tables to support arbitrarily complicated user-defined
sequences of restrictions and exceptions, with the ultimate in flexibility
being the facility to consult an external process which can implement
whatever logic is required: if you wish you can restrict some users to
sending mail on the third Tuesday after pay day only, if that is of any use
to you.  Administrators can also supply their own rejection messages, to
make it clear to senders why exactly their mail was rejected.
Unfortunately this flexibility has a cost: complexity in the logs
generated.  While it is easy to use \texttt{grep(1)} to determine the fate
of an individual email, following the progress of an email through Postfix
can be quite difficult.  The logs tend to follow a 90\%/10\% pattern: 90\%
of the time following the log entries is simple, but the other 10\% of the
logging requires 90\% of the code.

@@

This paper and the program it describes are part of a larger project to
optimise your Postfix restrictions, generate statistics and graphs, and
provide a platform on which new restrictions can be trialled and evaluated
to see if they are worth using in the fight against spam.  The program
parses Postfix logs and populates a database with the data gleamed from
those logs, providing a consistent and simple view of your logs which
future tools can utilise.  The program should be reasonably easily
extensible, so if logging to a database doesn't meet your requirements, you
can implement your own processing.

@@

The complexity and variation in Postfix's logs requires similar flexibility
in the parser; decoupling the parsing rules from the associated actions
allows the rules to be loaded from an external source without modifications
to the parser source code (significantly lowering the barrier to entry for
new users), and greatly simplifies both parser and rules.  The separation
also allows for a clear separation of functionality: rules handle low level
details of identifying a line, whereas the parser handles the higher level
details of following the path a mail takes through postfix, assembling the
required information, etc.

@@

The most substantive part of the parser is the parser al . . . . blah

@@

    \item If the mail is rejected, log all data, and finish with the mail.

    \item Follow the progress of the mail until it's either delivered or
        bounced, log all data, and finish with the mail.

@@

    \item The mail lacks a queueid until the mail has been accepted, so log
        lines must first be correlated by the smtpd pid, then transition to
        being correlated by the queueid.  This is relatively minor, but
        does require two versions of several functions,
        \texttt{per\_pid} and \texttt{per\_queueid}.

    \item Multiple independent mails may be delivered during one
        connection; this requires cloning the current data, so that
        subsequent mails won't trample over each other.  This must be done
        every time a mail is accepted, as it's impossible to tell in
        advance which connections will accept multiple mails.  Happily once
        the mail has been accepted log entries won't be correlated by pid
        for that mail any more, so there isn't any ambiguity about which
        mail a given log line belongs to.  One possible (as yet unseen)
        difficulty is distinguishing between rejects which belong to the
        current mail, and rejects which belong to previous, unaccepted,
        mails.

    \item The most difficult complication is that mails are not always
        delivered directly to a mailbox: sometimes they are aliased and
        need to be delivered to an address (or addresses) on another
        server.  When this occurs a child mail will be injected into the
        postfix queue, but without the explicit logging smtpd or sendmail
        injected mails have, so the source is not immediately discernible;
        from a strictly linear reading of the logs it \textit{usually}
        appears as if the child mail has appeared from thin air.
        Subsequently the parent mail will log the creation of the child
        mail:

        \texttt{3FF7C4317: to=<username@cs.tcd.ie>, relay=local, \newline 
        delay=0, status=sent (forwarded as 56F5B43FD)}

        Unfortunately while all log lines from an individual process appear
        in the logs in monotonic order, the order in which log lines from
        different processes are interleaved is subject to the vagaries of
        process scheduling.  In addition the first log line belonging to
        the child mail (the log line cited above properly belongs to the
        parent mail) is logged by qmgr, so depending on how busy qmgr (and
        the server generally) is, the parent's creation log line may appear
        before or after (generally after) the child's log line from qmgr.
        This has the effect that the parser cannot complain when it
        encounters a log line from qmgr for a previously unseen mail; it
        must mark the mail as potentially fake and subsequently clear the
        faked flag if and when the origin of the mail becomes clear.
        Obviously the parser could omit checking of where mails originate
        from, but I feel that it is better to require an explicit source,
        as bugs are more likely to be exposed.

@@

\includegraphics{logparser-flow-chart.ps}

\section{Limitations and improvements}

\section{Summary}

XXX TODO what happens to the original connection?? XXX

XXX TODO is it possible to distinguish between different rejected mails on
one connection??  XXX

>>>>

\title{Parsing Postfix log files}
\author{John Tobin \\ School of Computer Science and Statistics \\ 
Trinity College \\ Dublin 2 \\ Ireland \\ tobinjt@cs.tcd.ie}
\date{}
\maketitle

\begin{abstract}

    Parsing Postfix logs is much more difficult than it first appears, but
    it is possible.  This paper describes the process required, documenting
    the algorithm and difficulties encountered, with reference to an
    implementation which stores data gathered from the logs in an SQL
    database.

\end{abstract}

XXX WHAT IS THE CONCLUSION??? XXX

\newpage
\tableofcontents

\section{Introduction}

Most mail server administrators will have performed some basic processing
of Postfix logs at one time or another, whether it was to debug a problem,
explain to a user why mail is being rejected, or to check whether their new
anti-spam measures are working.  The more adventurous will have generated
some statistics to show how many hits each of their anti-spam measures have
gotten in the last week, and possibly even generated some graphs to clearly
illustrate the point to management or complaining users.\footnote{This was
the author's first real foray into processing Postfix logs.}  Very few will
have performed in-depth parsing and analysis of their logs, where the
parsing must correlate the log lines per-connection or per-queueid, rather
than processing lines independently.  One of the barriers to this type of
processing is the unstructured nature of Postfix's logs, where each logging
line was added on an ad hoc basis as a requirement was discovered or new
functionality was added.  Further complication arises the ability to have
rejection messages defined by the administrator, and every Real time Black
List returns a different message.  This paper documents the difficult
process of parsing Postfix logs, and presents a program which parses logs
and places the resulting data into a database for further processing by
other applications.

@@

Postfix is a highly configurable, high performance, secure and scalable
Mail Transport Agent.  It features extensive anti-spam restrictions,
allowing a mail administrator to deploy those restrictions which they judge
suitable for their needs, rather than a fixed set chosen by Postfix's
author.  These restrictions can be applied and combined on a per-host,
per-recipient or per-sender basis, allowing clients/recipients to be
whitelisted with respect to specific restrictions.  Postfix leverages
simple lookup tables to support arbitrarily complicated user-defined
sequences of restrictions and exceptions, with the ultimate in flexibility
being the facility to consult an external process which can implement
whatever logic is required: if you wish you can restrict some users to
sending mail on the third Tuesday after pay day only, if that is of any use
to you.  Administrators can also supply their own rejection messages, to
make it clear to senders why exactly their mail was rejected.
Unfortunately this flexibility has a cost: complexity in the logs
generated.  While it is easy to use \texttt{grep(1)} to determine the fate
of an individual email, following the progress of an email through Postfix
can be quite difficult.  The logs tend to follow a 90\%/10\% pattern: 90\%
of the time following the log entries is simple, but the other 10\% of the
logging requires 90\% of the code.\footnote{These numbers don't have a
solid scientific basis, they're based on gut feeling from writing the
software.}

@@

This paper and the program it describes are part of a larger project to
optimise your Postfix restrictions, generate statistics and graphs, and
provide a platform on which new restrictions can be trialled and evaluated
to see if they are worth using in the fight against spam.  The program
parses Postfix logs and populates a database with the data gleamed from
those logs, providing a consistent and simple view of your logs which
future tools can utilise.  The program should be reasonably easily
extensible, so if logging to a database doesn't meet requirements, one
can implement one's own processing.

\subsection{Parser background}

Before getting into detail about the parser, a brief overview is in order.
The parser is split into two parts: the parsing algorithm, and the rules
which are applied to the lines.

    \item [Rules] The first matching rule wins: no further rules are tried
        against that line,\footnote{With one exception, detailed later.}
        but there is a facility for the specifying the order of the rules,
        so that more specific rules can be matched first.  Rules are
        completely self-contained: there are no sub-rules, so the
        complexity is linear.  Rules are similar to the tokeniser/Lex part
        of a Lex/YACC parser.

    \item [Algorithm]

@@

The complexity and variation in Postfix's logs requires similar flexibility
in the parser; decoupling the parsing rules from the associated actions
allows the rules to be loaded from an external source without modifications
to the algorithm source code (significantly lowering the barrier to entry
for new users), and greatly simplifies both algorithm and rules.  The
separation also allows for a clear separation of functionality: rules
handle low level details of identifying a line, whereas the algorithm
handles the higher level details of following the path a mail takes through
postfix, assembling the required information, etc.

@@

While the rules are more than double (treble with 2.3.x rules?) the size of
the parser, the rules are quite simple, and each rule is very much
independent of its fellows.  The parser is significantly more complicated,
and highly interdependent.

@@

    \item If the mail is rejected, log all data and finish.

    \item Follow the progress of the mail until it's either delivered or
        bounced, log all data, and finish.

@@

    \item The mail lacks a queueid until it has been accepted, so log lines
        must first be correlated by the smtpd pid, then transition to being
        correlated by the queueid.  This is relatively minor, but does
        require two versions of several functions, \texttt{by\_pid} and
        \texttt{by\_queueid}.

    \item Multiple independent mails may be delivered during one
        connection; this requires cloning the current data, so that
        subsequent mails won't trample over each other.  This must be done
        every time a mail is accepted, as it's impossible to tell in
        advance which connections will accept multiple mails.  It is quite
        easy to overlook this complication because only a small minority of
        connections accept more than one mail. Happily once the mail has
        been accepted log entries won't be correlated by pid for that mail
        any more, so there isn't any ambiguity about which mail a given log
        line belongs to.  The original connection will be discarded when
        the client disconnects.  One possible (as yet unseen) difficulty is
        distinguishing between rejects which belong to the current mail,
        and rejects which belong to previous, unaccepted, mails.

    \item The most difficult complication is that mails are not always
        delivered directly to a mailbox (or program; there is very little
        difference between the two): sometimes they are aliased and
        need to be delivered to an address (or addresses) on another
        server.  When this occurs a child mail will be injected into the
        postfix queue, but without the explicit logging smtpd or sendmail
        injected mails have, so the source is not immediately discernible;
        from a strictly linear reading of the logs it \textit{usually}
        appears as if the child mail has appeared from thin air.
        Subsequently the parent mail will log the creation of the child
        mail:

        \texttt{3FF7C4317: to=<username@example.com>, relay=local, \newline 
        delay=0, status=sent (forwarded as 56F5B43FD)}

        Unfortunately while all log lines from an individual process appear
        in the logs in monotonic order, the order in which log lines from
        different processes are interleaved is subject to the vagaries of
        process scheduling.  In addition the first log line belonging to
        the child mail (the log line cited above properly belongs to the
        parent mail) is logged by qmgr, so depending on how busy qmgr (and
        the server generally) is, the parent's creation log line may appear
        before or after (usually after) the child's log line from
        qmgr.\footnote{Postfix is quite paranoid about mail delivery, an
        excellent characteristic for an MTA to posess, so it won't log that
        the child has been created until it is absolutely certain that the
        mail has been written to disk.  Because the parent then has to be
        informed that the mail has been committed, it's most likely that
        the qmgr will log first.}

        This has the effect that the parser cannot complain when it
        encounters a log line from qmgr for a previously unseen mail; it
        must mark the mail as potentially fake and subsequently clear the
        faked flag if and when the origin of the mail becomes clear.
        Obviously the parser could omit checking of where mails originate
        from, but I feel that it is better to require an explicit source,
        as bugs are more likely to be exposed.

@@

\includegraphics{logparser-flow-chart.ps}
\label{flow-chart}

\section{Limitations and possible improvements}

Every piece of software suffers from some limitations, and there is almost
always room for improvement.

\subsection{Limitations}

\begin{itemize}

    \item Each new Postfix release requires new rules to be written, to
        cope with the new messages.

    \item The program should save state at the end of each run, and reload
        it at the start of the next run, to properly cope with mails which
        span logs.  This might also require purging old mails once they've
        been in the state tables for too long.

    \item It appears that the hostname used in the HELO command is not
        logged if the mail is accepted.\footnote{Tested with Postfix
        2.2.10, this may possibly have changed in Postfix 2.3, or the
        upcoming 2.4.}

\end{itemize}

\subsection{Possible improvements}
\begin{itemize}

    \item A progress bar would be useful when run interactively, as the
        program takes roughly one minute per 10MB of
        logs.\footnote{Approximately 80\% of the run time is consumed by
        logging to the database.}  Obviously performance is entirely
        dependant on the machine the program is running on.

\end{itemize}

\section{Conclusion}
